The X-linked intellectual disability gene KDM5C is a sex-biased brake against germline programs in somatic lineages

### 2024-09-27 submission/manuscript.Rmd
remove extraneous print
shigeki edits

@@

Mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs) with unknown mechanisms. Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms and consequences of their dysregulation are largely unknown. Here, we examine how the X-linked intellectual disability gene lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3), contributes to tissue identity. We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key human neurological phenotypes, aberrantly express many liver, muscle, ovary, and testis genes within the amygdala and hippocampus. Gonad-enriched genes misexpressed in the _Kdm5c_-KO brain are unique to germ cells, indicating an erosion of the soma-germline boundary. Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_. Characterizing germline gene misexpression in males and female mutants revealed germline gene repression is sexually dimorphic, with female EpiLCs requiring a higher dose of KDM5C to maintain germline gene suppression. Using a comprehensive list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation (CpGme) during ESC to EpiLC differentiation. However, late-stage spermatogenesis genes devoid of promoter CGIs can become expressed in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors. Together, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a brake against runaway activation of germline developmental programs in somatic lineages.

@@

Distinguishing between germ cells and somatic cells is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] that occurs during early embryogenesis in many metazoans[@lehmannGermlineStemCells2012]. In mammals, chromatin regulators are crucial for decommissioning germline genes during the transition from na&#239;ve to primed pluripotency. Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone H3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in post-implantation epiblast cells[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021]. The contribution of KDM5C to this process remains unclear. Additionally, studies on germline gene repression have primarily been conducted in males and focused on select marker genes important for early germ cell development, given the lack of a comprehensive list for germline-enriched genes. Therefore, it is unknown if the mechanism of repression differs between sexes or for different classes of germline genes, e.g. meiotic versus spermatid differentiation genes.

@@

Previous RNA sequencing (RNA-seq) of the adult male _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes unique to the testis[@scandagliaLossKdm5cCauses2017]. It is currently unknown if the testis is the only tissue type misexpressed in the _Kdm5c_-KO brain. We thus systematically tested whether other tissue-specific genes are misexpressed in the male brain with constitutive knockout of _Kdm5c_ (_Kdm5c^-/y^_, 5CKO)[@vallianatosMutuallySuppressiveRoles2020] by using a published list of mouse tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017].

@@

`r print(tiss_list)`

@@

Although not consistent across brain regions, we also found significant enrichment of genes biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 2)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 2)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test) (Figure 1A-B). _Apolipoprotein C-I (Apoc1)_, a lipoprotein metabolism and transport gene, is among the liver-enriched DEG derepressed in both the hippocampus and amygdala[@roulandRoleApolipoproteinC12022] and its brain overexpression has been implicated in Alzheimer's disease[@leducAPOECholesterolHomeostasis2010] (Figure 1E).

@@

_Kdm5c_-KO brain expresses testicular germline genes[@scandagliaLossKdm5cCauses2017], however the testis also contains somatic cells that support hormone production and germline functions. To determine if _Kdm5c_-KO results in ectopic expression of somatic testicular genes, we first evaluated the known functions of testicular DEGs through gene onotology. We found _Kdm5c_-KO testis-enriched DEGs had high enrichment of germline-relevant ontologies, including spermatid development (GO: 0007286, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (GO: 0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A, Supplementary Table 1).

@@

KDM5C binds to the promoters of several germline genes in embryonic stem cells (ESCs) but not in neurons[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023]. However, the lack of a comprehensive list of germline-enriched genes prohibited genome-wide characterization of KDM5C binding at germline gene promoters. Thus, it is unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.

@@

Previous work found two germline gene promoters have a marked reduction in DNA CpG methylation (CpGme) in the adult _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017]. Since histone 3 lysine 4 di- and trimethylation (H3K4me2/3) impede _de novo_ CpGme[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's removal of H3K4me2/3 may be required to suppress germline genes. However, KDM5C's catalytic activity was recently shown to be dispensible for suppressing _Dazl_ in undifferentiated ESCs[@guptaGenomewideScreenReveals2023]. To reconcile these observations, we hypothesized KDM5C erases H3K4me2/3 to promote the initial placement of CpGme at germline gene promoters in EpiLCs.

@@

Promoters that showed the most robust loss of CpGme in _Kdm5c_-KO exEpiLCs (lowest q-values) harbored CGIs (Figure 6K). CGI promoters, but not CGI-free promoters, had a significant reduction in CpGme with KDM5C loss as a whole (Figure 6L) (Non-CGI promoters p = 0.0846, CGI promoters p = 0.0081, Mann-Whitney U test). Significantly hypomethylated promoters included germline genes consistently dysregulated across multiple _Kdm5c_-KO RNA-seq datasets[@scandagliaLossKdm5cCauses2017], such as _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) and _Naa11_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "Naa11", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "Naa11", "qvalue"),3)`) (Figure 6M). Surprisingly, we observed only a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6N). Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs in EpiLCs to promote CpGme at germline gene promoters. Furthermore, this suggests while KDM5C's cataltyic activity is required for the repression of some germline genes, CpGme can be placed at others even with elevated H3K4me2/3 around the TSS.

# Discussion
In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity. We first characterized tissue-enriched genes expressed within the mouse _Kdm5c_-KO brain and identified substantial derepression of testis, liver, muscle, and ovary-enriched genes. Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to the germline and absent in somatic cells. _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the adult brain primarily expresses genes important for late spermatogenesis. We demonstrated that although sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs have heightened germline gene de-repression with KDM5C loss. Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via indirect mechanisms, such as activation through ectopic RFX transcription factors. Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation to promote long-term germline gene silencing through DNA methylation at CpG islands. Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary. By systematically characterizing KDM5C's role in germline gene repression, we unveiled unique mechansisms governing the misexpression of distinct germline gene classes within somatic lineages. Ultimately, these data provide molecular footholds which can be exploited to test the overarching contribution of ectopic germline gene expression upon neurodevelopment.

By comparing _Kdm5c_ mutant males and females, we revealed germline gene supression is sexually dimorphic. Female EpiLCs are more severely impacted by loss of KDM5C-mediated germline gene suppression, yet this difference is not due to the large number of germline genes on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. Heightened germline gene misexpression in females may be related to females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, heterozygous knockout females (_Kdm5c^-/+^_) also had over double the number of germline DEGs than hemizygous knockout males (_Kdm5c^-/y^_), even though their expression of KDM5C should be roughly equivalent to that of wild-type males (_Kdm5c^+/y^_) . Males could partially compensate for KDM5C's loss via the Y-chromosome homolog, KDM5D. However, KDM5D exhibits weaker demethylase activity than KDM5C[@iwaseXLinkedMentalRetardation2007] and has not been reported to regulate germline gene expression. Nevertheless, these results demonstrate germline gene silencing mechanims differ between males and females, which warrants further study to elucidate the biological ramifications and underlying mechanisms.

We found KDM5C is largely dispensable for promoting normal gene expression during development, yet is critical for suppressing ectopic developmental programs. It is important to note that while we highlighted KDM5C's repression of germline genes, some germline-enriched genes like _Dazl_ are also expressed at the 2-cell stage and in the inner cell mass/na&#239;ve ESCs for their role in pluripotency and self-renewal[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023]. These "self-renewal" germline genes are then silenced during ESC differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021]. We found that while _Kdm5c_-KO EpiLCs express _Dazl_, they did not express 2-cell-specific genes like _Zscan4c_. These data suggest the 2-cell-like state reported in _Kdm5c_-KO ESCs[@guptaGenomewideScreenReveals2023] likely reflects KDM5C's primary role in germline gene repression. Germline gene misexpression in _Kdm5c_-KO EpiLCs may indicate they are differentiating into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011; @magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014]. Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which blocks EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. In addition to unimpaired EpiLC differentiation, _Kdm5c_-KO gross brain morphology is overall normal[@iwaseMouseModelXlinked2016] and hardly any brain-specific genes were significantly dysregulated in the amygdala and hippocampus. Thus, ectopic germline gene expression occurs in conjunction with overall proper somatic differentiation in _Kdm5c_-KO animals.

Our work provides novel insight into the cross-talk between H3K4me2/3 and CpGme, which are generally mutually exclusive[@weberDistributionSilencingPotential2007]. In EpiLCs, loss of KDM5C binding at a subset of germline gene promoters, e.g. _D1Pas1_, strongly impaired promoter CGI methylation and resulted in their long-lasting de-repression into adulthood. Removal of H3K4me2/3 at CGIs is a plausible mechanism for KDM5C-mediated germline gene suppression[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023], given H3K4me2/3 can oppose DNMT3 activity[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015]. However, emerging work indicates many histone-modifying enzymes have non-cataltyic functions that influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020]. Indeed, KDM5C's catalytic activity was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023]. In our study, _Dazl_'s promoter still gained CpGme in _Kdm5c_-KO exEpiLCs, even with elevated H3K4me2. _Dazl_ and a few other germline genes employ multiple repressive mechanisms to facilitate CpGme, such as DNMT3A/B recruitment via E2F6 and MGA[@pohlersRoleE2F6Restriction2005; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. This suggests alternative silencing mechansims are sufficient to recruit DNMT3s to some germline CGIs, while others may require KDM5C-mediated H3K4me removal to overcome promoter CGI escape from CpGme[@longProtectionCpGIslands2016; @weberDistributionSilencingPotential2007]. These results also suggest the requirement for KDM5C's catalytic activity can change depending upon the locus and developmental stage, even for the same class of genes. However, further experiments are required to determine if catalytically inactive KDM5C can suppress germline genes at later developmental stages.

By generating a comprehensive list of mouse germline-enriched genes, we revealed distinct derepressive mechanisms governing early versus late-stage germline programs. Previous work on germline gene silencing has focused on genes with promoter CGIs[@borgelTargetsDynamicsPromoter2010; @weberDistributionSilencingPotential2007], and indeed the majority of KDM5C targets in EpiLCs were germ cell identity genes harboring CGIs. However, over 70% of germline-enriched gene promoters lacked CGIs, including the many KDM5C-unbound germline genes that are de-repressed in _Kdm5c_-KO cells. CGI-free, KDM5C-unbound germline genes were primarily late-stage spermatogenesis genes and significantly enriched for RFX2 binding sites, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016]. These data suggest that once activated during early embryogenesis, drivers of germline gene expression like _Rfx2_, _Stra8_, and _Dazl_ turn on downstream germline programs, ultimately culminating in the expression of spermiogenesis genes in the adult _Kdm5c_-KO brain. Therefore, we propose KDM5C is recruited via promoter CGIs to act as a brake against runaway activation of germline-specific programs.

The above work provides the mechanistic foundation for KDM5C-mediated repression of tissue and germline-specific genes. However, the contribution of these ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown. In addition to germline genes, we also identified significant enrichment of muscle and liver-enriched transcripts within the _Kdm5c_-KO brain. Intriguingly, select liver and muscle-enriched DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022]. _APOC1_ dysregulation is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010] and overexpression of _Apoc1_ in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008]. KDM5C may therefore be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_.

>>>>

Correspondence should be addressed to K. Bonefas and S. Iwase (siwase@umich.edu)

@@

Mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs). Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms and consequences of their dysregulation are largely unknown. Here, we examine how the X-linked intellectual disability gene lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3), contributes to tissue identity. We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key human neurological phenotypes, aberrantly express many liver, muscle, ovary, and testis genes within the amygdala and hippocampus. Gonad-enriched genes misexpressed in the _Kdm5c_-KO brain are unique to germ cells, indicating an erosion of the soma-germline boundary. Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_. Characterizing germline gene misexpression in males and female mutants revealed germline gene repression is sexually dimorphic, with female EpiLCs requiring a higher dose of KDM5C to maintain germline gene suppression. Using a comprehensive list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation during ESC to EpiLC differentiation. However, late-stage spermatogenesis genes devoid of promoter CGIs can become expressed in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors. Together, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a brake against runaway activation of germline developmental programs in somatic lineages.

@@

Distinguishing between germ cells and somatic cells is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] that occurs during early embryogenesis in many metazoans[@lehmannGermlineStemCells2012]. In mammals, chromatin regulators are crucial for decommissioning germline genes during the transition from na&#239;ve to primed pluripotency. Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone H3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in post-implantation epiblast cells[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021]. The contribution of KDM5C to this process remains unclear. Additionally, studies on germline gene repression have primarily been conducted in males and focused on select marker genes, given the lack of a comprehensive list for germline-enriched genes. Therefore, it is unknown if the mechanism of repression differs between sexes or for different classes of germline genes, e.g. meiotic versus spermatid differentiation genes.

@@

Previous RNA sequencing (RNA-seq) of the adult male _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes unique to the testis[@scandagliaLossKdm5cCauses2017]. It is currently unknown if the testis is the only tissue type misexpressed in the _Kdm5c_-KO brain. We thus systematically tested whether other tissue-specific genes are misexpressed in the male brain with constitutive knockout of _Kdm5c_ (_Kdm5c^-/y^_, 5CKO in figures)[@vallianatosMutuallySuppressiveRoles2020] by using a published list of mouse tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017].

@@

@@

Although not consistent across brain regions, we also found significant enrichment of genes biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 2)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 2)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test) (Figure 1A-B). These include _Apolipoprotein C-I (Apoc1)_, a lipoprotein metabolism and transport gene[@roulandRoleApolipoproteinC12022] (Figure 1E,see Discussion).

@@

_Kdm5c_-KO brain expresses testicular germline genes[@scandagliaLossKdm5cCauses2017] (Figure 1), however the testis also contains somatic cells that support hormone production and germline functions. To determine if _Kdm5c_-KO results in ectopic expression of testicular somatic genes, we first evaluated the known functions of testicular DEGs through gene onotology. We found _Kdm5c_-KO testis-enriched DEGs had high enrichment of germline-relevant ontologies, including spermatid development (GO: 0007286, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (GO: 0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A, Supplementary Table 1).

@@

KDM5C binds to the promoters of several germline genes in embryonic stem cells (ESCs) but not in neurons[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023]. However, due to the lack of a comprehensive list of germline-enriched genes, it is unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.

@@

Previous work found two germline gene promoters have a marked reduction in DNA CpG methylation (CpGme) in the adult _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017]. Since histone H3K4me2/3 impede _de novo_ CpGme[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's removal of H3K4me2/3 may be required to suppress germline genes. However, KDM5C's catalytic activity was recently shown to be dispensible for suppressing _Dazl_ in undifferentiated ESCs[@guptaGenomewideScreenReveals2023]. To reconcile these observations, we hypothesized KDM5C erases H3K4me2/3 to promote the initial placement of CpGme at germline gene promoters in EpiLCs.

@@

Promoters that showed the most robust loss of CpGme in _Kdm5c_-KO exEpiLCs (lowest q-values) harbored CGIs (Figure 6K). CGI promoters, but not CGI-free promoters, had a significant reduction in CpGme with KDM5C loss as a whole (Figure 6L) (Non-CGI promoters p = 0.0846, CGI promoters p = 0.0081, Mann-Whitney U test). Significantly hypomethylated promoters included germline genes consistently dysregulated across multiple _Kdm5c_-KO RNA-seq datasets[@scandagliaLossKdm5cCauses2017], such as _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) and _Naa11_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "Naa11", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "Naa11", "qvalue"),3)`) (Figure 6M). Unexpectedly, we observed only a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6N). Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs in EpiLCs to promote CpGme at those promoters. Furthermore, our data suggest while KDM5C's cataltyic activity is required for the repression of some germline genes, CpGme can be placed at others even with elevated H3K4me2/3 around the TSS.

# Discussion
In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity. We first characterized tissue-enriched genes expressed within the mouse _Kdm5c_-KO brain and identified substantial derepression of testis, liver, muscle, and ovary-enriched genes. Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to the germline and absent in somatic cells. _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the adult brain primarily expresses genes important for late spermatogenesis. We demonstrated that although sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs have heightened germline gene de-repression with KDM5C loss. Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via indirect mechanisms, such as activation through ectopic RFX transcription factors. Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation to promote long-term germline gene silencing through CGI DNA methylation. Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary. By systematically characterizing KDM5C's role in germline gene repression, we unveiled distinct mechansisms governing the misexpression of distinct germline gene classes in somatic lineages. Ultimately, these data provide molecular footholds which can be exploited to test the overarching contribution of ectopic germline gene expression upon neurodevelopment.

By comparing _Kdm5c_ mutant males and females, we revealed germline gene supression is sexually dimorphic. Female EpiLCs are more severely impacted by loss of KDM5C-mediated germline gene suppression, yet this difference is not due to the large number of germline genes on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. Heightened germline gene misexpression in females may be related to females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, heterozygous knockout females (_Kdm5c^-/+^_) also had over double the number of germline DEGs than hemizygous knockout males (_Kdm5c^-/y^_), even though their expression of KDM5C should be roughly equivalent to that of wild-type males (_Kdm5c^+/y^_) . Males could partially compensate for KDM5C's loss via the Y-chromosome homolog, KDM5D[@iwaseXLinkedMentalRetardation2007]. However, KDM5D has not been reported to regulate germline gene expression. Nevertheless, these results demonstrate germline gene silencing mechanims differ between males and females, which warrants further study to elucidate the biological ramifications and underlying mechanisms.

We found KDM5C is largely dispensable for promoting normal gene expression during development, yet is critical for suppressing ectopic developmental programs. While some germline genes, such as _Dazl_, are also expressed in the 2-cell stage, inner cell mass, and na&#239;ve ESCs, they are silenced in epiblast stem cells/EpiLCs[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023; @mochizukiRepressionGermlineGenes2021]. Our data suggest the 2-cell-like state reported in _Kdm5c_-KO ESCs[@guptaGenomewideScreenReveals2023] likely reflects KDM5C's primary role in germline gene repression (Figure 3). Germline gene misexpression in _Kdm5c_-KO EpiLCs may indicate they are differentiating into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011; @magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014]. Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which blocks EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. In addition to unimpaired EpiLC differentiation, _Kdm5c_-KO gross brain morphology is overall normal[@iwaseMouseModelXlinked2016] and hardly any brain-specific genes were significantly dysregulated in the amygdala and hippocampus. Thus, ectopic germline gene expression occurs in conjunction with overall proper somatic differentiation in _Kdm5c_-KO animals.

Our work provides novel insight into the cross-talk between H3K4me2/3 and CpGme, which are generally mutually exclusive[@weberDistributionSilencingPotential2007]. In EpiLCs, loss of KDM5C binding at a subset of germline gene promoters, e.g. _D1Pas1_, strongly impaired promoter CGI methylation and resulted in their long-lasting de-repression into adulthood. Removal of H3K4me2/3 at CGIs is a plausible mechanism for KDM5C-mediated germline gene suppression[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023], given H3K4me2/3 repell DNMT3 activity[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015]. However, emerging work indicates many histone-modifying enzymes have non-cataltyic functions that influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020]. Indeed, KDM5C's catalytic activity was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023]. In our study, _Dazl_'s promoter still gained CpGme in _Kdm5c_-KO exEpiLCs, even with elevated H3K4me2. _Dazl_ and a few other germline genes employ multiple repressive mechanisms to facilitate CpGme, such as DNMT3A/B recruitment via E2F6 and MGA[@pohlersRoleE2F6Restriction2005; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. Thus, while some germline CGIs require KDM5C-mediated H3K4me removal to overcome promoter CGI escape from CpGme[@longProtectionCpGIslands2016; @weberDistributionSilencingPotential2007], others do not. These results also suggest the requirement for KDM5C's catalytic activity can change depending upon the locus and developmental stage. Further experiments are required to determine if catalytically inactive KDM5C can suppress germline genes at later developmental stages.

By generating a comprehensive list of mouse germline-enriched genes, we revealed distinct derepressive mechanisms governing early versus late-stage germline programs. Previous work on germline gene silencing has focused on genes with promoter CGIs[@borgelTargetsDynamicsPromoter2010; @weberDistributionSilencingPotential2007], and indeed the majority of KDM5C targets in EpiLCs were germ cell identity genes harboring CGIs. However, over 70% of germline-enriched gene promoters lacked CGIs, including the many KDM5C-unbound germline genes that are de-repressed in _Kdm5c_-KO cells. CGI-free, KDM5C-unbound germline genes were primarily late-stage spermatogenesis genes and significantly enriched for RFX2 binding sites, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016]. These data suggest that once activated during early embryogenesis, drivers of germline gene expression like _Rfx2_, _Stra8_, and _Dazl_ turn on downstream germline programs, ultimately culminating in the expression of spermiogenesis genes in the adult _Kdm5c_-KO brain. Therefore, we propose KDM5C is recruited via promoter CGIs to act as a brake against runaway activation of germline-specific programs. Future studies should adress how KDM5C is targeted to CGIs.

The above work provides the mechanistic foundation for KDM5C-mediated repression of tissue and germline-specific genes. However, the contribution of these ectopic, tissue-specific genes towards neurological impairments is still unknown. In addition to germline genes, we also identified significant enrichment of muscle and liver-enriched transcripts within the _Kdm5c_-KO brain. Intriguingly, select liver and muscle-enriched DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022]. _APOC1_ dysregulation is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010] and overexpression of _Apoc1_ in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008]. KDM5C may therefore be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_.

Given that germline genes have no known functions within the brain, their impact upon neruodevelopment is currently unknown. In _C. elegans_, somatic misexpression of germline genes via loss of _Retinoblastoma (Rb)_ homologs results in enhanced piRNA signaling and ectopic P granule formation in neurons[@wangSomaticMisexpressionGermline2005; @wuRepressionGermlineRNAi2012]. Ectopic testicular germline transcripts have also been observed in a variety of cancers, including brain turmors in _Drosophila_ and mammals[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017] and shown to promote cancer progression[@janicEctopicExpressionGermline2010; @ghafouri-fardExpressionCancerTestis2012; @ninBiologyCancerTestisAntigens2023]. Intriguingly, mouse models and human cells for other chromatin-linked NDDs also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014], such as mutations in DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, and methyl-CpG -binding protein 2 (MECP2). Recently, the transcription factor ZMYM2 (ZNF198), whose mutation causes an NDD (OMIM #619522), was also shown to repress germline genes by promoting H3K4me removal and CpGme[@graham-paquinZMYM2EssentialMethylation2023]. Thus, KDM5C is among a growing cohort of neurodevelopmental disorders with erosion of the germline-soma boundary. Further research is required to determine the impact of these germline genes upon neuronal functions and the extent to which this phenomenon occurs in humans.
